---
title: "Archetype cluster clinical variables"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (steps["archetypal_analysis"] == 0) {
  # If analysis was skipped, load saved results from previous run
	load("/home/.../archetypal_results.Rdata")
  
  # Output directory
	out_dir2 <- "/home/.../results/"
	if(arch_opts["package"] == 2) {
		use_archetypal <- T
		use_archetypes <- F
	} else {
		use_archetypal <- F
		use_archetypes <- T
	}
	
	do_adj <- ifelse(adj_dens == 1, T, F)
	arch_colour <- "Paired"

}	

```


Leo Dahl  
2020-04-07

This is step number `r step_count` in a workflow containing the steps  
`r step_list`.  
Main script started on `r starting_time`, this step started on `r Sys.time()`.  
Output files are generated in  
`r out_dir2`.

Normalisations that were performed:  
**`r names(norm_opts[norm_opts == 1])`**

Proteomics data adjusted for:  
**`r names(which(linnorm_opts == 1))`**

Package used:  
`r if(use_archetypes){"**archetypes**"}else if(use_archetypal){"**archetypal**"}`


<details><summary>Click for details on options</summary>
This document was generated with the following options:

```{r print options}

print(all_opts)

```

</details>

<details><summary>Click for details on packages</summary>
This document was generated using the following package versions

```{r}

print(R.Version()$version.string)
print_packversions()

```

</details>


## Clinical parameters

Several clinical parameters are investigated to see whether they drive the archetypes somehow.

Continuous variables are tested using the Wilcoxon rank sum test pairwise between clusters. The data is plotted in density plots and boxplots. 

Discrete variables are tested using the Fisher's exact test pairwise between clusters and plotted in bar charts (counts and proportions). 

#### Age

```{r Age, warning=F}

plot_results(arch_df, groupby="cluster", clin_var="age")

```


#### BMI

```{r BMI, warning=F}

bmiplot <- plot_results(arch_df, "cluster", "bmi", printplots=F)
bmiplot$boxplot + labs(y = "BMI")

```


#### Mammographic breast density

```{r Mammographic breast densities, warning=F}

adplot <- plot_results(arch_df, "cluster", "area.density", printplots=F)
adplot$boxplot + labs(y = "Dense area (cm2)")

# BMI-adjusted densities
if (do_adj) { 
  adadjplot <- plot_results(arch_df, "cluster", "area.dens.adj", printplots=F)
  adadjplot$boxplot + labs(y = "BMI-adjusted dense area (cm2)")
}

```


#### Case Control status

```{r Case control status, warning=F}

plot_results(arch_df, "cluster", "class")

```

#### PRS 

Polygenic risk scores (PRS)  

```{r PRS plot}

# PRS plot with cluster 1 vs group consisting of the other clusters
prs_clust1 <- arch_df %>%
  select(cluster, PRS) %>%
  mutate(cluster = replace(cluster, cluster != 1, "Other")) %>%
  filter(!is.na(PRS))

plot_results(prs_clust1, "cluster", "PRS")

```

***

