---
title: "Linear normalisation"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
params: 
  step_vec: 0
  step_nr: 0
  start_time: 0
  norms: 0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.duplicate.label = "allow")
```

Leo Dahl  
2020-02-25

This is step number `r params$step_nr` in a workflow containing the steps  
`r params$step_vec`.  
Main script started on `r params$start_time`, this step started on `r Sys.time()`.  
Output files are generated in  
`r "home/.../results/"`.


<details><summary>Click for details on options</summary>
This document was generated with the following options:

```{r print options}

print(all_opts)

```

</details>

<details><summary>Click for details on packages</summary>
This document was generated using the following package versions

```{r}

print(R.Version()$version.string)
print_packversions()

```

</details>


# Normalisation by linear regression model

So far, the data has been normalised for techincal variation using **`r names(params$norms[params$norms == 1])`**. Now the data will be adjusted for other factors such as age and study entry date. 

#### Removing flagged antibodies

Start by removing the antibodies that were flagged in the antibody quality control. The data contains `r length(which(binfo1$alt.assay.name == "A1"))`, 
`r length(which(binfo1$alt.assay.name == "A2"))`, 
`r length(which(binfo1$alt.assay.name == "A3"))` and 
`r length(which(binfo1$alt.assay.name == "A4"))` 
antibodies in assays A1, A2, A3 and A4, respectively, before removal. This corresponds to `r length(unique_prot)` unique proteins targeted by `r length(unique_ab)` unique antibodies (note that A1 and A2 contain the same antibodies). 

Plotting a UMAP projection of the MFI values (only for cases, controls and doubles) and colouring by different parameters results in the following plots

```{r UMAP scatter plots before removal, warning=F}

umap_bf_removal <- plot_umap(umap_casecontrol(mfi, sinfo),
          sinfo[sinfo$Class == "Case" | sinfo$Class == "Control" | sinfo$Class == "Other", ], T, F)
umap_bf_removal$age
umap_bf_removal$bmi
umap_bf_removal$entrydate

```


```{r}
# New binder info without A2 and flagged antibodies, also those lacking ENSG ID since they are either controls or Abs that have failed homology tests 
binfo_all <- binfo1
binfo <- binfo1[binfo1$alt.assay.name != "A2" & binfo1$abqc_fail == 0 &
                  binfo1$ENSG.ID != "", ]
```

A look at the CV% and IQR of binders

```{r techincal and biological cv}

# Look at techincal and biological CV of antibodies
cv <- function(invec) 100*sqrt(exp(sd(invec)**2) - 1)

b.cv <- apply(mfi[sinfo %>% filter(Class %in% c("Case", "Control", "Other") & replicate != "2") %>% pull(new.names), 
									binfo$alt.ab.name], 2, cv)
t.cv <- apply(mfi[sinfo %>% filter(Class == "Pool") %>% pull(new.names), 
									binfo$alt.ab.name], 2, cv)

cvdf <- data.frame("binder"=names(b.cv), 
									 "assay"=binfo[names(b.cv), "alt.assay.name"], 
									 "b.CV"=b.cv, 
									 "t.CV"=t.cv[names(b.cv)],
									 "b.CV.rankall"=rank(b.cv),
									 "t.CV.rankall"=rank(t.cv[names(b.cv)]))

cvdf$t.CV.rankassay = cvdf$b.CV.rankassay <- NA
for (aname in c("A1", "A2", "A3", "A4")) {
	cvdf[cvdf$assay == aname, "b.CV.rankassay"] <- rank(cvdf[cvdf$assay == aname, "b.CV"])
	cvdf[cvdf$assay == aname, "t.CV.rankassay"] <- rank(cvdf[cvdf$assay == aname, "t.CV"])
}

# Plot CV for technical variation and IQR for biological variation
tcv_plt <- ggplot(cvdf) + 
	geom_point(aes(x = t.CV.rankall, y = t.CV)) + 
	geom_hline(yintercept = median(cvdf$t.CV), linetype = "dotted", size = 1) + 
	labs(x = "Antibody, CV order", y = "Technical CV%", caption = paste0("Median t.CV: ", median(cvdf$t.CV) %>% signif(2))) + 
  theme_classic()
tcv_plt

ab_iqr <- apply(mfi[sinfo %>% 
											filter(Class %in% c("Case", "Control", "Other")) %>% 
											pull(new.names),
										binfo$alt.ab.name], 2, IQR)

iqrdf <- data.frame(ab_iqr, order=rank(ab_iqr))

iqr_plt <- ggplot(iqrdf) + 
	geom_point(aes(x = order, y = ab_iqr)) + 
	geom_hline(yintercept = median(iqrdf$ab_iqr), linetype = "dotted", size = 1) + 
	labs(x = "Antibody, IQR order", y = "IQR", 
			 caption = paste0("Median IQR: ", median(iqrdf$ab_iqr) %>% signif(2))) + 
  theme_classic()
iqr_plt

```

Euclidean distance and correlation between pairs of samples (double, replicate, random)

```{r euclidean distance and correlation of samples, message=F}

# Make function that computes euclidean distance or correlation between a number of random sample pairs, not very general function
rnd_eucl_corr <- function(out_len, method="distance") {
	out_vec <- vector("numeric", out_len)
	
	if (method == "distance") {
		
		for (i in 1:out_len) {
			rnd_samp <- sinfo %>% 
				filter(Class %in% c("Case", "Control") & replicate != "2") %>% 
				slice(sample(sinfo %>% 
										 	filter(Class %in% c("Case", "Control") & replicate != "2") %>% 
										 	nrow(), 2)) %>%
				pull(new.names)
			out_vec[i] <- dist(mfi[rnd_samp, binfo$alt.ab.name])
		}
		
	} else if (method == "correlation") {
		
		for (i in 1:out_len) {
			rnd_samp <- sinfo %>% 
				filter(Class %in% c("Case", "Control") & replicate != "2") %>% 
				slice(sample(sinfo %>% 
										 	filter(Class %in% c("Case", "Control") & replicate != "2") %>% 
										 	nrow(), 2)) %>%
				pull(new.names)
			out_vec[i] <- cor(t(mfi[rnd_samp[1], binfo$alt.ab.name]),
												t(mfi[rnd_samp[2], binfo$alt.ab.name]), 
												method="spearman")
		}
	}
	return(out_vec)
}

set.seed(123)
# Euclidean distances between doubles
dbl_eucl <- sapply(unique(sinfo %>% 
														filter(Class == "Other" & replicate != "2") %>%
														pull(Tube.label)), 
									 function(x) {
									 	
									 	dist(mfi[sinfo %>% 
									 					 	filter(Tube.label == x & replicate != "2") %>% 
									 					 	pull(new.names), binfo$alt.ab.name])
									 
									 	})

rnd_eucl_dbl <- rnd_eucl_corr(length(dbl_eucl))

# Euclidean distances between replicated samples
repl_eucl <- sapply(unique(sinfo %>% 
													 	filter(replicate %in% c("1", "2")) %>% 
													 	pull(Sample.id)), 
										function(x) {
											
											dist(mfi[sinfo %>% 
															 	filter(Sample.id == x) %>% 
															 	pull(new.names),
															 binfo$alt.ab.name])
											
										})

rnd_eucl_repl <- rnd_eucl_corr(length(repl_eucl))

# Correlation between doubles
dbl_corr <- sapply(unique(sinfo %>% 
														filter(Class == "Other" & replicate != "2") %>%
														pull(Tube.label)), 
									 function(x) {
									 	
									 	dbl_samp <- sinfo %>% 
									 		filter(Tube.label == x & replicate != "2") %>%
									 		pull(new.names)
									 	cor(t(mfi[dbl_samp[1], binfo$alt.ab.name]),
									 			t(mfi[dbl_samp[2], binfo$alt.ab.name]),
									 			method="spearman")
									 })

rnd_corr_dbl <- rnd_eucl_corr(length(dbl_corr), "correlation")

# Correlation between replicated samples
repl_corr <- sapply(unique(sinfo %>% 
													 	filter(replicate %in% c("1", "2")) %>%
													 	pull(Sample.id)),
										function(x) {
											
											repl_samp <- sinfo %>%
												filter(Sample.id == x) %>%
												pull(new.names)
											cor(t(mfi[repl_samp[1], binfo$alt.ab.name]), 
													t(mfi[repl_samp[2], binfo$alt.ab.name]), 
													method="spearman")
											
										})

rnd_corr_repl <- rnd_eucl_corr(length(repl_corr), "correlation")

# For plotting
dbl_stats <- data.frame(dbl_eucl, dbl_corr, rnd_eucl_dbl, rnd_corr_dbl) %>% melt()
repl_stats <- data.frame(repl_eucl, repl_corr, rnd_eucl_repl, rnd_corr_repl) %>% melt()

# Save plots for QC summary plot in manuscript_figures.Rmd
dbl_eucl_dens <- ggplot(dbl_stats %>% filter(variable %in% c("dbl_eucl", "rnd_eucl_dbl"))) + 
	geom_density(aes(x = value, y = ..scaled.., colour = variable), size = 2) + 
	labs(x = "Euclidean distance", y = "Density", colour = "Sample") + 
	scale_colour_hue(labels = c("dbl_eucl"="Double", "rnd_eucl_dbl"="Random")) + 
	xlim(2, 15) + 
  theme_classic()
repl_eucl_dens <- ggplot(repl_stats %>% filter(variable %in% c("repl_eucl", "rnd_eucl_repl"))) +
	geom_density(aes(x = value, y = ..scaled.., colour = variable), size = 2) +
	labs(x = "Euclidean distance", y = "Density", colour = "Sample") + 
	scale_colour_hue(labels = c("repl_eucl"="Replicate", "rnd_eucl_repl"="Random")) +
	xlim(2, 15) +
  theme_classic()

dbl_eucl_dens
repl_eucl_dens

```


```{r Remove bad abs and replicate samples and assay}
# Keep mfi data frame with replicates for use in analysis prep script
mfi_w_repl <- mfi[
  rownames(sinfo[sinfo$replicate == 1 | sinfo$replicate == 2, ]),
  c("sample.antibody", binfo$alt.ab.name)]
mfi_w_repl$replicate <- sinfo[rownames(mfi_w_repl), "replicate"]

# Remove replicate samples and replicate assay, as well as empty samples and pools
sinfo_w_repl <- sinfo
sinfo <- sinfo[(sinfo$replicate == 0 | sinfo$replicate == 1)
               & (is.na(sinfo$entry.date) == F & sinfo$entry.date != ""), ]

# MFIs and sample info of replicates to use for estimating their residuals
repl_sinfo <- sinfo_w_repl[which(sinfo_w_repl$replicate == 2 & sinfo_w_repl$Class != "Pool"), ]
repl_mfi <- mfi_w_repl[rownames(repl_sinfo), -1]

mfi <- mfi[sinfo$new.names, 
           c("sample.antibody", binfo$alt.ab.name)]

unique_prot <- get_unique_prot(binfo)
unique_ab <- unique(binfo$Binder.name)

```


```{r Remove doubles before analysis}

# Remove doubles before performing linear normalisation
# Keep the first sample of doubles (or whichever if same age at sampling)
doub <- unique(sinfo[which(sinfo$Class == "Other"), "Tube.label"])
double_indices <- c()
for (d in doub) {
  samps <- which(sinfo$Tube.label == d)
  ages <- sinfo[samps, "Sampling.age"]
  
  if (ages[1] == ages[2]) {
    double_indices <- append(double_indices, samps[1]) 
  } else {
    double_indices <- append(double_indices, samps[which(ages == max(ages))])
  }
}

mfi_w_doubles <- mfi
mfi <- mfi[-double_indices, ]

sinfo_w_doubles <- sinfo
sinfo <- sinfo[-double_indices, ]

# MFIs and sample info of doubles to use for estimating their residuals
double_sinfo <- sinfo_w_doubles[double_indices, ]
double_sinfo <- double_sinfo[which(is.na(double_sinfo$bmi) == F), ] # Remove those missing BMI information
double_mfi <- mfi_w_doubles[rownames(double_sinfo), -1]

### Save MFI for later regression with proteins of interest and clinical variables
mfi_regr <- mfi

```


After removing the flagged antibodies, as well as all control antibodies and the antibodies missing binder info, the assays contain 
`r length(which(binfo1$alt.assay.name == "A1" & binfo1$abqc_fail == 0))` (A1 after removing union of flagged antibodies in A1 & A2), 
`r length(which(binfo1$alt.assay.name == "A3" & binfo1$abqc_fail == 0))` (A3) and 
`r length(which(binfo1$alt.assay.name == "A4" & binfo1$abqc_fail == 0))` (A4) 
antibodies. In total there are `r length(unique_ab)` unique antibodies that bind `r length(unique_prot)` unique proteins. 

Plotting UMAP projections again gives these plots

```{r UMAP scatter plots after removal, warning=F}

# Keep the plots to merge with others for reports
umap_after_removal <- plot_umap(umap_casecontrol(mfi, sinfo), 
          sinfo[sinfo$Class == "Case" | sinfo$Class == "Control" | sinfo$Class == "Other", ],
          T, printall=F)

# Print some here
umap_after_removal$age
umap_after_removal$bmi
umap_after_removal$entrydate

```

```{r logreg and clogit options, message=F, warning=F}

do_logreg <- ifelse(linnorm_opts["logreg"] == 1, T, F)

do_clogit <- ifelse(linnorm_opts["clogit"] == 1, T, F)

# If stuff here
if (do_logreg & do_clogit) {
	render(input = paste0(script_dir, "logreg_and_clogit.Rmd"), 
				 output_file = paste0(out_dir, Sys.Date(), "_logreg_and_clogit.html"))
} else if (do_logreg & !do_clogit) {
	render(input = paste0(script_dir, "logreg_and_clogit.Rmd"),
				 output_file = paste0(out_dir, Sys.Date(), "_logreg.html"))
} else if (!do_logreg & do_clogit) {
	render(input = paste0(script_dir, "logreg_and_clogit.Rmd"), 
				 output_file = paste0(out_dir, Sys.Date(), "_clogit.html"))
}

```


#### Normalisation

The data is normalised for age, BMI and entry date by generating a linear regression model for each protein where age and entry date are the independent variables. The acquired residuals are used as the protein measurement "values". 

```{r Linear regression normalisation, eval=T}

# Loop to make linear models for each protein, to normalise for age and entry date

# Get variables to adjust for, make formula for linear model
vars <- names(which(linnorm_opts[-which(names(linnorm_opts) %in% c("logreg", "clogit"))] == 1))
linfun <- as.formula(
  paste("values", paste0(vars, collapse=" + "), sep=" ~ ")
)

# To save p-values (why not just make data frame to begin with? whatever)
age_date_lm_pval <- vector("list", (ncol(mfi) - 1))
idx <- 1


# Initialize empty data frames to fill with new values
mfi_residuals <- data.frame(matrix(nrow=nrow(mfi),
                                   ncol=ncol(mfi),
                                   dimnames=list(rownames(mfi),
                                                 colnames(mfi))
                                   )
                            )
mfi_residuals[, 1] <- mfi[, 1]

repl_residuals <- data.frame(matrix(nrow=nrow(repl_mfi),
                                    ncol=(ncol(repl_mfi)-1)))
repl_residuals[, 1] <- repl_mfi[, 1]
rownames(repl_residuals) <- rownames(repl_mfi)
colnames(repl_residuals) <- colnames(repl_mfi)[1:(ncol(repl_mfi)-1)]

double_residuals <- data.frame(matrix(nrow=nrow(double_mfi),
                                      ncol=ncol(double_mfi),
                                      dimnames=list(rownames(double_mfi),
                                                    colnames(double_mfi))))

# Remove rows missing BMI values if adjusting for BMI
if (linnorm_opts["bmi"] == 1) {
  mfi_residuals <- mfi_residuals[-which(is.na(sinfo$bmi)), ]
  sinfo <- sinfo[-which(is.na(sinfo$bmi)), ]
}

for (ab in colnames(mfi)[-1]) {
  linreg_input <- data.frame("sample"=rownames(mfi),
                             "values"=mfi[, ab],
                             "age"=sinfo[rownames(mfi), "Sampling.age"],
                             "entry.date"=as.Date(sinfo[rownames(mfi),
                                                        "entry.date"]),
                             "bmi"=sinfo[rownames(mfi), "bmi"]
                             )
  
  age_date_lm <- lm(linfun, linreg_input)
  
  # Store p-vals
  age_date_lm_pval[[idx]] <- summary(age_date_lm)$coefficients[, "Pr(>|t|)"]
  
  mfi_residuals[, ab] <- age_date_lm$residuals
  
  # Get residuals for replicates and doubles by predicting their fitted values
  repl_predict_input <- data.frame("values"=repl_mfi[, ab],
                                   "age"=repl_sinfo[rownames(repl_mfi), "Sampling.age"],
                                   "entry.date"=as.Date(
                                     repl_sinfo[rownames(repl_mfi), "entry.date"]),
                                   "bmi"=repl_sinfo[rownames(repl_mfi), "bmi"])
  rownames(repl_predict_input) <- rownames(repl_mfi)
  
  repl_lmfit <- predict(age_date_lm, repl_predict_input)
  repl_residuals[, ab] <- repl_predict_input$values - repl_lmfit
  
  double_predict_input <- data.frame("values"=double_mfi[, ab],
                                     "age"=double_sinfo[rownames(double_mfi), "Sampling.age"], 
                                     "entry.date"=as.Date(
                                       double_sinfo[rownames(double_mfi),"entry.date"]),
                                     "bmi"=double_sinfo[rownames(double_mfi), "bmi"])
  rownames(double_predict_input) <- rownames(double_mfi)
  
  double_lmfit <- predict(age_date_lm, double_predict_input)
  double_residuals[, ab] <- double_predict_input$values - double_lmfit
  
  # Increment index for list
  idx <- idx + 1
}


mfi <- mfi_residuals

adpval <- data.frame(matrix(unlist(age_date_lm_pval),
                            ncol=(length(vars) + 1),
                            byrow=T))
colnames(adpval) <- names(age_date_lm_pval[[1]])

# Plot p-value distribution for varibales in linear regression
for (norm_opt in vars) {
  print(paste0(norm_opt, ": Number of genes with p-value < 0.05"))
  print(adpval %>% filter(get(norm_opt) < 0.05) %>% nrow())
}



```

Doing UMAP and plotting after normalisation results in the following plots.

```{r UMAP after normalisation, warning=F, eval=T}

umap_after_adj <- plot_umap(
  umap_casecontrol(mfi_residuals, sinfo), sinfo, T, printall=F)

# merge plots for age, plots for samp date and plots for bmi
age_adj_plots <- (umap_after_removal$age + labs(title = "Age, not adjusted")) +
  (umap_after_adj$age + labs(title = "Age, adjusted")) + 
  plot_layout(guides="collect")

entrydate_adj_plots <- (umap_after_removal$entrydate + labs(title = "Entry date, not adjusted")) +
  (umap_after_adj$entrydate + labs(title = "Entry date, adjusted")) + 
  plot_layout(guides="collect")

bmi_adj_plots <- (umap_after_removal$bmi + labs(title = "BMI, not adjusted")) +
  (umap_after_adj$bmi + labs(title = "BMI, adjusted")) + 
  plot_layout(guides="collect")

```


```{r compare age and bmi and entry date before and after adjustment, eval=T}

(umap_after_removal$age + labs(title = "Not adjusted for age")) + 
  (umap_after_adj$age + labs(title = "Adjusted for age")) + 
  plot_layout(guides = "collect")

(umap_after_removal$bmi + labs(title = "Not adjusted for BMI")) + 
  (umap_after_adj$bmi + labs(title = "Adjusted for BMI")) + 
  plot_layout(guides = "collect")

(umap_after_removal$entrydate + labs(title = "Not adjusted for entry date")) + 
  (umap_after_adj$entrydate + labs(title = "Adjusted for entry date")) + 
  plot_layout(guides = "collect")

```


```{r}

# Change linnorm_opts to not include the logreg and clogit options, for later output. The original options are still stored in the all_opts variable 
linnorm_opts <- linnorm_opts[-which(names(linnorm_opts) %in% c("logreg", "clogit"))]


```





