---
title: "Logistic Regression and Conditional Logistic Regression"
author: "Leo Dahl"
date: "2020-11-25"
output: 
  html_document: 
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Main script started on `r starting_time`, this script started on `r Sys.time()`.  
Output files are generated in  
`r "home/.../results/"`.

<details><summary>Click for details on options</summary>
This document was generated with the following options:

```{r print options}

print(all_opts)

```

</details>


`r if(do_logreg){"#### Logistic regression"}`

`r if(do_logreg){"Logistic regression is performed to evaluate whether any protein levels are significantly tied to case control status. A general logistic regression model is made with protein level, age, BMI and entry date as the predictors, to be able to assess the protein level effect independently of the other factors. "}`

```{r case control logistic regression, eval=do_logreg}

# Initialise data frame for storing p-values and coefficients for each antibody
cc_pcoeff <- data.frame(matrix(nrow=3,
                               ncol=(ncol(mfi) - 1)))
colnames(cc_pcoeff) <- colnames(mfi)[-1]
rownames(cc_pcoeff) <- c("coefficient", "pvalue", "adjusted.pval")

for (ab in colnames(mfi)[-1]) {
  # Make input data frame containing case control status, protein levels, age, bmi and  entrydate in each loop
  logreg_input <- data.frame("protein"=mfi[sinfo[which(sinfo$Class %in% c("Case", "Control")), "new.names"], ab],
                             "class"=sinfo[which(sinfo$Class %in% c("Case", "Control")), "Class"],
                             "age"=sinfo[which(sinfo$Class %in% c("Case", "Control")), "Sampling.age"],
                             "bmi"=sinfo[which(sinfo$Class %in% c("Case", "Control")), "bmi"], 
                             "entrydate"=sinfo[which(sinfo$Class %in% c("Case", "Control")), "entry.date"])
  
  # Need numeric arguments, convert case control status into 0 (control) and 1 (case), and entry date into numbers (using unclass function)
  logreg_input[which(logreg_input$class == "Control"), "class"] <- 0
  logreg_input[which(logreg_input$class == "Case"), "class"] <- 1
  logreg_input$class <- as.numeric(logreg_input$class)
  logreg_input$entrydate <- logreg_input$entrydate %>%
    as.Date() %>%
    unclass()
  
  cc_glm <- glm(class ~ protein + age + bmi + entrydate, data=logreg_input, family = binomial)
  
  cc_pcoeff[1:2, ab] <- c(summary(cc_glm)$coefficients["protein", "Estimate"],
                          summary(cc_glm)$coefficients["protein", "Pr(>|t|)"])
  
}

# Adjust the p-values for multiple test correction usign the false discovery rate method ("FDR")
cc_pcoeff["adjusted.pval", ] <- p.adjust(cc_pcoeff["pvalue", ], method="fdr")
cc_pcoeff <- as.data.frame(t(cc_pcoeff))

```

`r if(do_logreg){"Number of adjusted p-values (FDR) below 0.05 is "}`
`r if(do_logreg){length(which(cc_pcoeff$adjusted.pval < 0.05))}`

`r if(do_logreg){"P-value distributions (unadjusted and adjusted):"}`

```{r logreg pval distribution, eval=do_logreg}

ggplot(cc_pcoeff) + geom_histogram(aes(x = pvalue)) + geom_vline(xintercept = 0.05, linetype = "dashed") + labs(caption = "Dashed line: p = 0.05")

ggplot(cc_pcoeff) + geom_histogram(aes(x = adjusted.pval)) + geom_vline(xintercept = 0.05, linetype = "dashed") + labs(caption = "Dashed line: adj.p = 0.05")

```


`r if(do_clogit){"#### Conditional logistic regression"}`


`r if(do_clogit){"This code makes a conditional logistic model for each binder in the data to investigate the impact of the case control status on the protein levels. All models have the matching cases and controls (match ID) taken into consideration. For independent variables the models use \n\n - Model 1: BMI and entry date \n\n - Model 2: model 1 + area-based breast density, menopausal status, and HRT treatment \n\n - Model 3: model 2 + parity, smoking, and alcohol consumtion \n\n ***"}`


```{r clogit prep, eval=do_clogit}

sinfo_clogit <- sinfo %>% filter(Class == "Case" | Class == "Control") 

# Merge HRT ever variable from survey and prescription registry
# If entry from registry is missing or is zero (not taken), use value from survey, as registry only goes back to around 2005

rownames(sinfo_clogit) <- sinfo_clogit$Tube.label
sinfo_clogit$hrt_ever_merge <- sapply(sinfo_clogit$Tube.label, function(id) {
	survey_hrt <- sinfo_clogit[id, "hrt.ever"]
	registry_hrt <- hrt_statin_info[id, "hrt_ever"]
	
	hrt_out <- ifelse((is.na(registry_hrt) & !is.na(survey_hrt)) | 
											(registry_hrt == 0 & !is.na(survey_hrt)), 
										survey_hrt, registry_hrt)
	
	return(hrt_out)
})

sinfo_clogit <- sinfo_clogit %>%
	select(Sample.id, Tube.label, Class, bmi, entry.date, matchid, 
				 area.dens, postmenop, hrt_ever_merge, 
				 smoke_pack_year, alcohol_g_week, parity)

sinfo_clogit$entry.date <- as.Date(sinfo_clogit$entry.date)


trim_matchsamp <- function(indf) {
  ### Function for trimming input data frame to not contain samples that are missing any matched samples ("lonely samples") or two matched controls without any matched case
  
  # Get number of samples for each matchid
  matchtbl <- table(indf$matchid)
  
  # Get matchid for lonely samples
  onesamp <- names(matchtbl[matchtbl == 1])
  
  # Get matchid for samples where only two controls remain
  twosamp_id <- names(matchtbl[matchtbl == 2])
  twosamp_case <- indf %>% 
    filter(matchid %in% twosamp_id) %>%
    filter(Class == 1) %>%
    pull(matchid)
  twosamp_nocase <- setdiff(twosamp_id, twosamp_case)
  
  # Remove lonely samples and pairs without cases
  outdf <- indf %>%
    filter(!matchid %in% onesamp & !matchid %in% twosamp_nocase)
  
  return(outdf)
}

```

```{r clogit no imputation, message=F, warning=F, eval=do_clogit}
library(survival)

# clogit procedure is performed without controls with diagnoses 6 and 16 years before sampling (both same matchid)
matchid_remove <- sinfo %>% 
  filter(Class == "Control" & prevalent_entry == "1") %>% 
  pull(matchid) %>% 
  unique()

# For storing p-values
clog_res <- data.frame(matrix(nrow = (ncol(mfi) - 1), 
                              ncol = 7))
rownames(clog_res) <- colnames(mfi)[-1]
colnames(clog_res) <- c("binder", paste0("p_value_model", 1:3), paste0("effect_size_model", 1:3))
clog_res$binder <- rownames(clog_res)


# Conditional logistic regression 
for (ab in colnames(mfi)[-1]) {
	f_model1 <- paste0("Class ~ ", ab, " + bmi + entry.date + strata(matchid)")
	f_model2 <- paste0(f_model1, " + area.dens + postmenop + hrt_ever_merge")
	f_model3 <- paste0(f_model2, " + smoke_pack_year + alcohol_g_week + parity")
	
	clog_in <- cbind(mfi[sinfo_clogit$Sample.id, ab], sinfo_clogit) %>% filter(!matchid %in% matchid_remove)
	colnames(clog_in)[1] <- ab
	clog_in$Class <- ifelse(clog_in$Class == "Case", 1, 0) # Class variable should be numeric or logical for clogit function to work

	clog_in$matchid <- as.numeric(clog_in$matchid)
	# Remove samples with missing values for clinical variables, remove match IDs where there is no matched case left or only one sample left
	clog_in <- clog_in %>% filter(!is.na(bmi) & !is.na(entry.date))
	clog_in <- trim_matchsamp(clog_in)
	
	clog_model1 <- clogit(as.formula(f_model1), clog_in) %>% summary()
	
	clog_in <- clog_in %>% filter(!is.na(area.dens) & !is.na(postmenop) & !is.na(hrt_ever_merge))
	clog_in <- trim_matchsamp(clog_in)
	
	clog_model2 <- clogit(as.formula(f_model2), clog_in) %>% summary()
	
	clog_in <- clog_in %>% filter(!is.na(smoke_pack_year) & !is.na(alcohol_g_week) & !is.na(parity))
	clog_in <- trim_matchsamp(clog_in)
	
	clog_model3 <- clogit(as.formula(f_model3), clog_in) %>% summary()
	
	clog_res[ab, "p_value_model1"] <- clog_model1$coefficients[ab, "Pr(>|z|)"]
	clog_res[ab, "p_value_model2"] <- clog_model2$coefficients[ab, "Pr(>|z|)"]
	clog_res[ab, "p_value_model3"] <- clog_model3$coefficients[ab, "Pr(>|z|)"]
	
	clog_res[ab, "effect_size_model1"] <- clog_model1$coefficients[ab, "coef"]
	clog_res[ab, "effect_size_model2"] <- clog_model2$coefficients[ab, "coef"]
	clog_res[ab, "effect_size_model3"] <- clog_model3$coefficients[ab, "coef"]
	
}

# -log10 of p-values
clog_res[, paste0("logP_model", 1:3)] <- -log10(clog_res[, paste0("p_value_model", 1:3)])
# FDR-corrected p-values
clog_res[, paste0("fdr_model", 1:3)] <- apply(clog_res[, paste0("p_value_model", 1:3)], 2, function(x) p.adjust(x, method="fdr"))
```


```{r, eval=F, fig.show="hold", out.width="50%"}
for (mod in 1:3) {
  
  plt_title <- paste0("Conditional logistic regression, model ", mod)
	# Without labels
	volc_plt <-
		clog_res %>%
		mutate("p_colour" = ifelse(get(paste0("logP_model", mod)) > -log10(0.05),
														 ifelse(get(paste0("logP_model", mod)) > 2, "2", "1"),
														 "0")) %>%
		ggplot(aes_string(x = paste0("effect_size_model", mod),
											y = paste0("logP_model", mod))) +
		geom_point(aes(colour = p_colour)) +
		geom_hline(yintercept = 2, linetype = "dotted") +
		geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
		scale_colour_manual(values = c("2"="red", "1"="black", "0"="gray")) +
	  labs(title = plt_title) + 
		theme_classic() +
		theme(legend.position = "none")

	# With labels
	volc_plt_lbl <-
		clog_res %>%
		mutate("p_colour" = ifelse(get(paste0("logP_model", mod)) > -log10(0.05),
														 ifelse(get(paste0("logP_model", mod)) > 2, "2", "1"),
														 "0")) %>%
		ggplot(aes_string(x = paste0("effect_size_model", mod),
											y = paste0("logP_model", mod))) +
		geom_point(aes(colour = p_colour)) +
		geom_hline(yintercept = 2, linetype = "dotted") +
		geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
		scale_colour_manual(values = c("2"="red", "1"="black", "0"="gray")) +
		geom_label_repel(data = clog_res %>%
										 	mutate("binder" = ifelse(get(paste0("logP_model", mod)) > -log10(0.05),
										 													 rownames(clog_res), "")),
										 aes(label = binder),
										 size = 2.5) +
	  labs(title = plt_title) + 
		theme_classic() +
		theme(legend.position = "none")

	print(volc_plt)
	print(volc_plt_lbl)

}
```

```{r, eval=F}

for (i in 1:3) {
  qq(clog_res[, paste0("p_value_model", i)], 
     main=paste0("Model ", i))
}

```

```{r, eval=do_clogit}

print("Stats for FDR")
clog_res %>% summarise(FDR_min_model1 = min(fdr_model1),
											FDR_max_model1 = max(fdr_model1),
											FDR_median_model1 = median(fdr_model1)) 
clog_res %>% summarise(FDR_min_model2 = min(fdr_model2),
											FDR_max_model2 = max(fdr_model2),
											FDR_median_model2 = median(fdr_model2))
clog_res %>% summarise(FDR_min_model3 = min(fdr_model3), 
											FDR_max_model3 = max(fdr_model3), 
											FDR_median_model3 = median(fdr_model3))

clog_res %>% count(p_value_model1 < 0.05)
clog_res %>% count(p_value_model2 < 0.05)
clog_res %>% count(p_value_model3 < 0.05)

print("Number of proteins with p < 0.05 in one or more models: ")
clog_res %>%
	filter(p_value_model1 < 0.05 | p_value_model2 < 0.05 | p_value_model3 < 0.05) %>%
	nrow()

```

```{r option to save results, eval=T}

if (do_clogit) {
  save(clog_res, 
       file = paste0(out_dir, "../results_logreg_and_clogit/", Sys.Date(), "_clogit.Rdata"))
}

```


### Packages

<details><summary>Click for details on packages</summary>
This document was generated using the following package versions

```{r}

print(R.Version()$version.string)
print_packversions()

```

</details>


