---
title: "Make manuscript figures"
author: "Leo Dahl"
date: "2020-11-12"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style type="text/css">

h1 { 
  font-size: 25px;
  margin-top: 1.5cm;
  margin-bottom: 0.5cm;
}
h2 { 
  font-size: 18px;
  margin-top: 1cm;
  margin-bottom: 0.5cm;
}

</style>

<details><summary>Click for details on options</summary>
This document was generated with the following options:

```{r print options}

print(all_opts)

```

</details>


```{r extra package}

# For converting heatmap to ggplot object and make compatible with patchwork package
library(ggplotify)

```


Manuscript figures
==================

This script generates a few special figures that will be included in the manuscript. Uses objects from many scripts, run everything except 'variation_inspection' up to 'archetypal_analysis' and then also run 'new_clinvar' and 'enrichment_analysis'. Also tables and certain values that will be included in the text. 

***

PCA, cluster counts and heatmap
-------------------------------

```{r pca and counts, fig.width=10, fig.height=8}

library(ggplotify)
ggheat <- as.ggplot(phm2)

(pc12 + labs(tag = "A", colour = "Cluster") + theme_classic() + 
    theme(legend.title = element_text(size = 16),
          legend.text = element_text(size = 13)) + 
		pc13 + labs(tag = "B", colour = "Cluster") + theme_classic() + 
    theme(legend.title = element_text(size = 16),
          legend.text = element_text(size = 13))) / 
	(pc23 + labs(tag = "C", colour = "Cluster") + theme_classic() + 
	   theme(legend.title = element_text(size = 16),
	         legend.text = element_text(size = 13)) + 
	 	count_plot + theme_classic() + 
	   theme(legend.position = "none") + labs(tag = "D")) +
	plot_layout(guides = "collect") 

```


```{r heatmap and volcano, fig.width=9, fig.height=9}

plt_layout <- c(
	area(t=1,b=6,l=1,r=3),
	area(t=7,b=11,l=1,r=2)
)
(ggheat + labs(tag = "A")) + 
	(de_volcano + labs(tag = "B")) +
	plot_layout(design = plt_layout)

```


***

Age, BMI and BMI-adjusted dense area
------------------------------------

```{r age bmi dense area, fig.width=15, warning=FALSE}

(plot_results(arch_df, "cluster", "age", printplots=F, printp=F)$boxplot + 
 	labs(title = NULL, tag = "A")) + 
    (plot_results(arch_df, "cluster", "bmi", printplots=F, printp=F)$boxplot + 
     	labs(title = NULL, y = "BMI", tag = "B")) + 
    (plot_results(arch_df, "cluster", "area.dens.adj", printplots=F, printp=F)$boxplot +
     	labs(title = NULL, y = "BMI-adjusted dense area (cm^2)", tag = "C")) + 
    plot_layout(guides = "collect") & 
  theme_classic() &
	theme(axis.title = element_text(size = 20),
				axis.text = element_text(size = 16),
				legend.title = element_text(size = 20),
				legend.text = element_text(size = 16),
				plot.tag = element_text(size = 20)) 

```

***

Case-control status, HRT ever, HRT status, days between entry and HRT, cluster MHT and case/control status
----------------------------------------------------------------------------------------------------------

```{r case control hrt, fig.height=15, fig.width=14, warning=F, message=F}

# Contingency tables/heatmaps for number of people in the clusters based on MHT and case control status
clust_mht_class_counts <- arch_df %>%
  select(cluster, class, hrt_ever_merge2) %>%
  table() %>%
  as.data.frame() %>%
  group_by(cluster, class) %>%                        # Rescale data for fill colours in plot
  mutate(scale_clustclass = Freq/sum(Freq)) %>%       # Rescale individually per cluster and class
  mutate(percent_labs = paste0(Freq, "\n(", signif(100*scale_clustclass, 2), "%)"))

contingency_heatmap <- ggplot(clust_mht_class_counts, aes(x = 1, y = hrt_ever_merge2)) + 
  geom_tile(aes(fill = scale_clustclass)) + 
  scale_fill_gradient2(low = "white", mid = "white", high = "salmon") + 
  geom_text(aes(label = percent_labs), size = 6) + 
  facet_grid(rows = vars(class),
             cols = vars(cluster), 
             switch = "x") + 
  theme_classic() + 
  theme(panel.grid = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_text(size = 18), 
        axis.title = element_text(size = 20), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(), 
        legend.position = "none", 
        strip.text = element_text(size = 19),
        strip.background.x = element_blank(),
        panel.spacing.x = unit(0, "lines"),
        panel.spacing.y = unit(0, "lines")) + 
  labs(x = "Cluster", y = "Taken MHT ever") + 
  scale_y_discrete(labels = c("0"="No", "1"="Yes"))

casecontrol_hrt_plt_top <- ((plot_results(arch_df, "cluster", "class", printplots=F, printp=F)$rightplot + 
		labs(fill = "Class")) +
	(plot_results(arch_df, "cluster", "hrt_ever_merge2", printplots=F, printp=F)$rightplot + 
			labs(fill = "MHT taken \never") + 
			scale_fill_hue(labels = c("No", "Yes")))) /
	((plot_results(arch_df, "cluster", "hrt_status_merge2", printplots=F, printp=F)$rightplot + 
	 	labs(fill = "MHT status \nat entry") + 
	 	scale_fill_hue(labels = c("Never taken", "Taken before", "Taking"))) +
	(plot_results(arch_df, "cluster", "hrt_endentrydiff_merge", printplots=F, printp=F)$boxplot + 
	 	labs(title = NULL, y = "Years between study entry \nand MHT treatment"))) + 
  plot_annotation(tag_levels = "A") & 
  theme_classic() & 
	theme(axis.title = element_text(size = 20),
				axis.text = element_text(size = 18),
				legend.title = element_text(size = 20),
				legend.text = element_text(size = 16), 
				plot.tag = element_text(size = 20)) 

casecontrol_hrt_plt_top / contingency_heatmap + 
  theme(plot.tag = element_text(size = 20))


```


***

Supplementary figures and tables
================================

Normalisation and adjustment
----------------------------

Plots comparing before and after adjustment for age, bmi, entry date

```{r, fig.width=11, fig.height=16}

((umap_after_removal$age + labs(title = "Age\nUnadjusted")) +
   (umap_after_adj$age + labs(title = "Age\nAdjusted")) +
   plot_layout(guides = "collect")) /
  ((umap_after_removal$bmi + labs(title = "BMI\nUnadjusted")) +
     (umap_after_adj$bmi + labs(title = "BMI\nAdjusted")) +
     plot_layout(guides = "collect")) /
  ((umap_after_removal$entrydate + labs(title = "Entry date\nUnadjusted")) +
     (umap_after_adj$entrydate + labs(title = "Entry date\nAdjusted")) +
     plot_layout(guides = "collect")) +
  plot_annotation(tag_levels = "A") &
  theme_classic() &
  theme(axis.title = element_text(size = 15),
						legend.title = element_text(size = 15),
						legend.text = element_text(size = 13),
						plot.title = element_text(size = 15))

normplt_layout <- "
A#
A#
BB
BB
CC
CC
DD
DD
"
(ggplot(sinfo_t1 %>% select(hospital, UMAP1, UMAP2)) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = hospital), size = 1.7, alpha = 0.9) +
  labs(colour = "Sampling station") +
   scale_colour_manual(values = RColorBrewer::brewer.pal(8, "Paired")[c(2,4,6,8)],
                       labels = sort(unique(sinfo_t1$hospital)))) /
   ((umap_after_removal$age + labs(title = "Age\nUnadjusted")) +
      (umap_after_adj$age + labs(title = "Age\nAdjusted")) +
      plot_layout(guides = "collect")) /
   ((umap_after_removal$bmi + labs(title = "BMI\nUnadjusted")) +
      (umap_after_adj$bmi + labs(title = "BMI\nAdjusted")) +
      plot_layout(guides = "collect")) /
   ((umap_after_removal$entrydate + labs(title = "Entry date\nUnadjusted")) +
      (umap_after_adj$entrydate + labs(title = "Entry date\nAdjusted")) +
      plot_layout(guides = "collect")) +
  plot_annotation(tag_levels = "A") +
  plot_layout(design = normplt_layout) &
  theme_classic() &
  theme(axis.title = element_text(size = 15),
						legend.title = element_text(size = 15),
						legend.text = element_text(size = 13),
						plot.title = element_text(size = 15))

normplt_layout <- "
AAAA#
AAAA#
AAAA#
AAAA#
AAAA#
BBBBB
BBBBB
BBBBB
BBBBB
BBBBB
CCCCC
CCCCC
CCCCC
CCCCC
CCCCC
DDDDD
DDDDD
DDDDD
DDDDD
DDDDD
"

(ggplot(sinfo_t1 %>% select(hospital, UMAP1, UMAP2)) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = hospital), size = 1.7, alpha = 0.9) +
  labs(colour = "Sampling station") +
   scale_colour_manual(values = RColorBrewer::brewer.pal(8, "Paired")[c(2,4,6,8)],
                       labels = sort(unique(sinfo_t1$hospital)))) /
   ((umap_after_removal$age + labs(title = "Age\nUnadjusted")) +
      (umap_after_adj$age + labs(title = "Age\nAdjusted")) +
      plot_layout(guides = "collect")) /
   ((umap_after_removal$bmi + labs(title = "BMI\nUnadjusted")) +
      (umap_after_adj$bmi + labs(title = "BMI\nAdjusted")) +
      plot_layout(guides = "collect")) /
   ((umap_after_removal$entrydate + labs(title = "Entry date\nUnadjusted")) +
      (umap_after_adj$entrydate + labs(title = "Entry date\nAdjusted")) +
      plot_layout(guides = "collect")) +
  plot_annotation(tag_levels = "A") +
  plot_layout(design = normplt_layout) &
  theme_classic() &
  theme(axis.title = element_text(size = 15),
						legend.title = element_text(size = 15),
						legend.text = element_text(size = 13),
						plot.title = element_text(size = 15))

```

Correlation before and after normalisation
------------------------------------------

```{r}
cordens_nonorm + cordens_abspqn + 
	plot_layout(guides = "collect") &
  theme_classic()
```

Hospital UMAP
-------------

```{r, fig.height=5, fig.width=8.3}
ggplot(sinfo_t1 %>% select(hospital, UMAP1, UMAP2)) +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = hospital), size = 1.7, alpha = 0.9) +
  labs(colour = "Sampling station") +
   scale_colour_manual(values = RColorBrewer::brewer.pal(8, "Paired")[c(2,4,6,8)],
                       labels = sort(unique(sinfo_t1$hospital))) + 
  theme_classic() + 
  theme(axis.title = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 13),
        plot.title = element_text(size = 15))
```


Other QC plots
--------------

```{r, fig.width=12, fig.height=10}

plt_layout <- c(
	area(t=1, b=2, l=1, r=2),
	area(t=3, b=4, l=1, r=2),
	area(t=1, b=2, l=3, r=4),
	area(t=3, b=4, l=3, r=4),
	area(t=1, b=2, l=5, r=6)
)

dbl_eucl_dens + repl_eucl_dens + 
	tcv_plt + iqr_plt + 
	repl_assay_corrplt + 
	plot_layout(design = plt_layout) + 
	plot_annotation(tag_levels = "A") &
	theme(plot.tag = element_text(size = 12),
				axis.title = element_text(size = 15),
				axis.text = element_text(size = 12), 
				legend.title = element_text(size = 15),
				legend.text = element_text(size = 12),
				plot.caption = element_text(size = 12)) 

```

Hierarchical clustering dendrogram (outlier check)
--------------------------------------------------

```{r}
plot(outlier_clust)
```


Archetypal analysis related plots (also for supplementary)
----------------------------------------------------------

Scree plot from archetypal analysis (if find_optimal_kappas was run)

```{r}

if (find_opt_k) {
	plot(1:maxkappa, k_opt$bestfit_sse1$sse, type="o", 
			 xlab="Number of archetypes", ylab="SSE")
	
	# Horizontal and vertical lines for easier getting axis values at optimal k
	clip(0, k_opt$optimal_kappas, 0, k_opt$bestfit_sse1$sse[k_opt$optimal_kappas])
	abline(h=k_opt$bestfit_sse1$sse[k_opt$optimal_kappas], v=k_opt$optimal_kappas, lty="dashed", col="red", lwd=2.5)
}

```


Largest archetype coefficients of the different clusters

```{r, message=F, warning=F}

plot_results(arch_maxalphas, "cluster", "alpha_max", printplots=F)$boxplot + labs(title = NULL, y = "Max archetype coefficient") + theme_classic() + 
  theme(axis.text = element_text(size = 12), 
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 15))

```


Proportions of sample pairs ending up in the same cluster

```{r}

prop_plt + 
    scale_x_discrete(labels=paste0(as.character(levels(same_cluster_df$sample_type)), 
       paste0(" pair\nn=", same_cluster_df %>% count(sample_type) %>% pull(n)))) +
	theme(axis.title = element_text(size = 16),
				axis.text = element_text(size = 14),
				legend.title = element_blank(),
				legend.text = element_text(size = 14)) 

```

Difference between largest and second largest archetype coefficient, for predicted coefficients of replicates excluded from archetypal analysis and real coefficients of the used replicates ("originals"?).

```{r}

pred_diff + real_diff + 
	plot_annotation(tag_levels = "A") &
	theme(axis.title = element_text(size = 15),
				axis.text = element_text(size = 13),
				legend.title = element_text(size = 15),
				legend.text = element_text(size = 13), 
				plot.tag = element_text(size = 15))

```

***

Cluster statin use
------------------

```{r}

plt_layout <- c(
	area(t=1,b=4,l=1,r=10),
	area(t=1,b=4,l=11,r=20)
)

(plot_results(arch_df, "cluster", "statin_ever", printplots=F, printp=F)$rightplot +
  labs(fill = "Used statins \never") + 
  scale_fill_hue(labels = c("0"="No", "1"="Yes"))) + 
  (plot_results(arch_df %>% filter(statin_philicity != "None"), "cluster", "statin_philicity",
             printplots=F, printp=F)$rightplot +
  labs(fill = "Philicity of \nused statins")) + 
	plot_layout(design = plt_layout) + 
	plot_annotation(tag_levels = "A") & 
	theme(axis.title = element_text(size = 12),
				axis.text = element_text(size = 9),
				legend.title = element_text(size = 11),
				legend.text = element_text(size = 10), 
				plot.tag = element_text(size = 12)) &
  theme_classic()

```

Cluster Polygenic Risk Score (PRS)
----------------------------------

```{r, message=F}

plot_results(arch_df %>% mutate(cluster = ifelse(cluster == 1, cluster, "Other")), 
             "cluster", "PRS", printp=F, printplots=F)$boxplot + 
  theme_classic() + 
  plot_results(arch_df, "class", "PRS", printp=F, printplots=F)$boxplot + 
  theme_classic() + 
  scale_fill_manual(values = c("Case"="#F8766D",
                               "Control"="#00BFC4")) +
  plot_annotation(tag_levels = "A") &
  theme(axis.title = element_text(size = 20),
        axis.text = element_text(size = 18), 
        legend.position = "none", 
        plot.tag = element_text(size = 20)) &
  labs(title = NULL)

```


Linear association with proteins of interest
--------------------------------------------

```{r}

heatmap.2(t(as.matrix(hm2_df)),
          cellnote = t(as.matrix(hm2_p)),
          Colv = as.dendrogram(hm_clust), 
          Rowv = F,
          dendrogram = "column",
          ColSideColors = col_annot$colours,
          labRow = c("Adjusted\ndense area", 
                     "MHT status"),
          notecol = "black",
          trace = "none",
          key.title = "",
          key.xlab = "Regression \ncoefficient",
          col = cell_col, 
          colsep = 1:nrow(hm2_df),
          rowsep = 1:ncol(hm2_df), 
          sepwidth = c(0.001, 0.001),
          sepcolor = "grey",
          margins = c(6, 12),
          srtCol = 90,
          density.info = "none",
          cexRow = 1)

legend("topright", title = "DE direction", 
       legend = c("Upregulated", "Downregulated"),
       fill = c("lightskyblue1", "lightcoral"),
       cex = 0.7, box.lty = 0)

```

MFI levels of proteins of interest
----------------------------------

```{r, message=F, fig.width=15, fig.height=14}

# Look at PTCH1 and ZP4
ptchzp <- mfi %>% 
  select(binfo %>% filter(Gene.name %in% c("PTCH1", "ZP4")) %>% pull(alt.ab.name))
colnames(ptchzp) <- binfo[colnames(ptchzp), "Gene.name"]
ptchzp$sample <- rownames(ptchzp) 
ptchzp[, c("cluster", "class", "mht_ever")] <- 
  arch_df[match(arch_df$sample, ptchzp$sample), c("cluster", "class", "hrt_ever_merge2")]

ptchzp_list <- vector("list", 4)
names(ptchzp_list) <- paste0(rep(c("PTCH1", "ZP4"), each=2), 
                             c(".class", ".mht_ever"))

for (prot in c("PTCH1", "ZP4")) {
  for (cvar in c("class", "mht_ever")) {
    plt <- plot_results(ptchzp %>% filter(!is.na(mht_ever)), 
                        cvar, prot, printplots=F, printp=F)$boxplot + 
      labs(title = NULL, x = ifelse(cvar == "mht_ever", "MHT taken ever", "Class")) + 
      theme_classic() + 
      theme(axis.title = element_text(size = 20),
            axis.text = element_text(size = 18), 
            legend.position = "none", 
            plot.tag = element_text(size = 20))
    
    ptchzp_list[[paste0(prot, ".", cvar)]] <- plt
  }
}

wrap_plots(ptchzp_list) + plot_annotation(tag_levels = "A")

```


Tables of cohort demographics and archetype cluster characterstics
------------------------------------------------------------------

Sample demographics tables (without and with doubles)

```{r samp table}

# table strata and labels defined in get_data.Rmd
table1(tbl1_strata, tbl1_labels)
table1(tbl1_d_strata, tbl1_labels)

# Use the term MHT instead of HRT
tbl1_labels <- list(
  variables=list(Sampling.age="Age",
                 bmi="BMI",
                 hospital="Sampling station",
                 postmenop="Menopausal status", 
                 area.dens="Dense area (cm2)",
  							 hrt_status_merge="MHT status",
  							 statin_status="Statin status",
  							 smoke_pack_year="Smoking (packs per year)",
  							 alcohol_g_week="Alcohol intake (g per week)", 
  							 parity="Ever given birth", 
                 er_merge="ER status",
                 pr_merge="PR status",
                 her2_merge="HER2 status",
  							 inv_merge="Invasiveness",
                 tsize_merge="Tumour size",
                 n.recode="Lymph node metastasis",
                 nhg_merge="Nottingham Histologic Grade"
                 )
)

cont_var <- c("Sampling.age", "bmi", "area.dens", "smoke_pack_year", "alcohol_g_week")
cat_var <- c("hospital", "postmenop", "hrt_status_merge", "statin_status", "parity")

# Wilcoxon test for continuous variables (not age since matched)
tbl1_p_cont <- sapply(cont_var, function(clinvar) {
	wilcox.test(sinfo_t1 %>% filter(Class == "Case") %>% pull(clinvar), 
							sinfo_t1 %>% filter(Class == "Control") %>% pull(clinvar))$p.value
}) 

print(tbl1_p_cont)

# Fisher test for categorical variables (not sampling location since matched)
tbl1_p_cat <- sapply(cat_var, simplify = F,function(clinvar) {
	# Get counts for contingency table
	casecounts <- sinfo_t1 %>% filter(Class == "Case") %>% pull(clinvar) %>% table()
	controlcounts <- sinfo_t1 %>% filter(Class == "Control") %>% pull(clinvar) %>% table()
	
	grp <- sinfo_t1 %>% filter(Class != "Double") %>% pull(clinvar) %>% as.character() %>% unique()
	grp <- grp[!is.na(grp)]
	
	cont_tbl <- data.frame(matrix(nrow=length(grp), 
																ncol=2))
	rownames(cont_tbl) <- grp
	colnames(cont_tbl) <- c("Case", "Control")
	
	cont_tbl[grp, "Case"] <- as.numeric(casecounts[grp])
	cont_tbl[grp, "Control"] <- as.numeric(controlcounts[grp])
	
	fisher.test(cont_tbl)$p.value
})

print(tbl1_p_cat)


# Make table that includes the p-values

a <- sinfo_t1 %>% filter(Class != "Double")
a$Class <- factor(a$Class, 
									levels=c("Case", "Control", "pval"), 
									labels=c("Case", "Control", "P-value"))

# Pretty much the same as the P-value example from the table1 vignette, but with the p-values taken from the previously calculated ones and with "-" for tumour characteristics of controls
rndr <- function(x, name, ...) {
    if (length(x) == 0 &
  		name %in% union(cont_var, cat_var)) {
      y <- a[, name]
      s <- rep("", length(render.default(x=y, name=name, ...)))
      if (name %in% cont_var) {
          p <- tbl1_p_cont[name]
      } else if (name %in% cat_var) {
          p <- tbl1_p_cat[[name]]
      } 
      s[2] <- sub("<", "&lt;", format.pval(p, digits=3, eps=0.001))
      s
    } else if (length(x) == 0 &
  		!(name %in% union(cont_var, cat_var))) {
    	""
    } else {
    	# Tumour characteristics
    	if (length(x) %in% c(nrow(a %>% filter(Class == "Control")),
    											 nrow(sinfo_t1 %>% filter(Class != "Double"))) &
    			name %in% c("er_merge", "pr_merge", "her2_merge", "inv_merge", "tsize_merge", "n.recode", "nhg_merge")) {
    		s <- rep("-", length(render.default(x=x, name=name, ...)))
    		s[1] <- ""
    		s
    	} else {
    		render.default(x=x, name=name, ...)	
    	}
    }
}

rndr.strat <- function(label, n, ...) {
    ifelse(n==0, label, render.strat.default(label, n, ...))
}

t1_strat <- c(list(Total=a), split(a, a$Class))

table1(t1_strat, tbl1_labels, render=rndr, render.strat=rndr.strat, overall="Total")

sinfo_t1 %>% filter(Class != "Double") %>% 
	group_by(Class) %>%
	summarise(minMD=min(area.dens, na.rm=T),
						maxMD=max(area.dens, na.rm=T)) %>%
	as.data.frame()

# Make one that includes doubles too
rndr.dbl <- function(x, name, ...) {
    if (length(x) %in% c(nrow(sinfo_t1 %>% filter(Class == "Control")),
                         nrow(sinfo_t1 %>% filter(Class == "Double")),
    										 nrow(sinfo_t1)) &
        name %in% c("er_merge", "pr_merge", "her2_merge", "inv_merge", "tsize_merge", "n.recode", "nhg_merge")) {
        s <- rep("-", length(render.default(x=x, name=name, ...)))
        s[1] <- ""
        s
    } else {
        render.default(x=x, name=name, ...)
    }
}

table1(tbl1_d_strata, tbl1_labels, render=rndr.dbl, overall="Total")

```


Archetype cluster table 

```{r arch table}

# Table of characteristics of archetype clusters
arch_tbl_df <- arch_df

# Change some variables into factors for proper labelling 
arch_tbl_df$hrt_status_merge2 <- factor(arch_tbl_df$hrt_status_merge2, 
																				levels=c("0", "1", "2"), 
																				labels=c("Never taken", "Taken before", "Taking at entry"))

arch_tbl_df$statin_status <- factor(arch_tbl_df$statin_status, 
																		levels=c("0", "1", "2"), 
																		labels=c("Never taken", "Taken before", "Taking at entry"))

arch_tbl_df$postmenop <- factor(arch_tbl_df$postmenop, 
																levels=c("0", "1"), 
																labels=c("Premenopausal", "Postmenopausal"))

arch_tbl_df$parity <- factor(arch_tbl_df$parity, 
														 levels=c("0", "1"),
														 labels=c("Never given birth", "Has given birth"))

arch_tbl_df$er_merge <- factor(arch_tbl_df$er_merge,
															 levels=c("0", "1"),
															 labels=c("Negative", "Positive"))

arch_tbl_df$pr_merge <- factor(arch_tbl_df$pr_merge, 
															 levels=c("0", "1"), 
															 labels=c("Negative", "Positive"))

arch_tbl_df$her2_merge <- factor(arch_tbl_df$her2_merge, 
																 levels=c("0", "1"), 
																 labels=c("Negative", "Positive"))
arch_tbl_df$inv_merge <- factor(arch_tbl_df$inv_merge, 
																levels=c("0", "1"), 
																labels=c("Cancer in situ", "Invasive"))

arch_tbl_df$n.new <- factor(arch_tbl_df$n.new, 
														levels=c("0", "1"), 
														labels=c("No", "Yes"))

arch_lbl1<- list(
	variables=list(class="Case control status", 
								 age="Age", 
								 bmi="BMI",
								 hrt_status_merge2="MHT status",
								 statin_status="Statin status",
								 postmenop="Menopausal status",
								 area.density="Dense area (cm2)", 
								 area.dens.adj="BMI- and age-adjusted dense area (cm2)",
								 smoke_pack_year="Smoking (packs per year)",
								 alcohol_g_week="Alcohol intake (g per week)",
								 parity="Ever given birth"), 
	groups=list("", "Archetype cluster")
)

arch_lbl2 <- list(
	variables=list(er_merge="ER status", 
								 pr_merge="PR status", 
								 her2_merge="HER2 status", 
								 inv_merge="Invasiveness",
								 tsize_merge="Tumour size (TNM T)",
								 n.new="Lymph node metastasis (TNM N)", 
								 NHG_merge="Nottingham Histologic Grade"),
	groups=list("", "Archetype cluster")
)

arch_str1 <- c(list(Total=arch_tbl_df), split(arch_tbl_df, arch_tbl_df$cluster))
arch_str2 <- c(list(Total=arch_tbl_df %>% filter(class == "Case")),
							 split(arch_tbl_df %>% filter(class == "Case"),
							 			 arch_tbl_df %>% filter(class == "Case") %>% pull(cluster)))

arch_tbl1 <- table1(arch_str1, arch_lbl1, groupspan=c(1, n_arch))
arch_tbl2 <- table1(arch_str2, arch_lbl2, groupspan=c(1, n_arch))

arch_tbl1
arch_tbl2

```


```{r save the data}

# Saving the data 
save(sinfo, sinfo_t1, binfo, arch, arch_df, mfi, file="/home/.../results/archetypal_results.Rdata")

```


<details><summary>Click for details on packages</summary>
This document was generated using the following package versions

```{r}

print(R.Version()$version.string)
print_packversions()

```

</details>
