---
title: "normalisation"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
params: 
  step_vec: 0
  step_nr: 0
  start_time: 0
  absPQN: 0
  ma96: 0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Leo Dahl  
2020-02-11

This is step number `r params$step_nr` in a workflow containing the steps  
`r params$step_vec`.  
Main script started on `r params$start_time`, this step started on `r Sys.time()`.  
Output files are generated in  
`r "home/.../results/"`.


<details><summary>Click for details on options</summary>
This document was generated with the following options:

```{r print options}

print(all_opts)

```

</details>

<details><summary>Click for details on packages</summary>
This document was generated using the following package versions

```{r}

print(R.Version()$version.string)
print_packversions()

```

</details>


# Normalisation

UMAP of the data before normalisation.

```{r some preparation}

# Triggers for displaying text and performing methods
if (params$absPQN != 0) {
  do_abspqn <- TRUE
} else {
  do_abspqn <- FALSE
}

if (params$ma96 != 0) {
  do_ma96 <- TRUE
} else {
  do_ma96 <- FALSE
}

# Log transform data
mfi[, -1] <- log(mfi[, -1])

for (assay in 1:length(mfi_list)) {
  mfi_list[[assay]] <- log(mfi_list[[assay]])
}

# Remove empty wells and antibodies that failed CET since not used at all from here on (?)
# But keep old ones for safety
mfi_w_empty <- mfi
sinfo_w_empty <- sinfo
binfo1_w_cetfail <- binfo1

mfi <- mfi[sinfo$Class != "Empty", ]
sinfo <- sinfo[sinfo$Class != "Empty", ]
binfo1 <- binfo1[binfo1$cet.fail == 0, ]


```


```{r umap and density before normalisation, warning=F}

# Perform and plot UMAP before normalisation
umap_nonorm <- umap_casecontrol(mfi, sinfo)

plot_umap(umap_nonorm, sinfo[sinfo$Class == "Case" | sinfo$Class == "Control" | sinfo$Class == "Other", ], T, F)$class

# Plot correlation density before normalisation
cordens_nonorm <- cor_dens(mfi, sinfo, xrange=c(0.85, 1), yrange=c(0, 340), plotname="No normalisation")
cordens_nonorm

```


`r if(do_abspqn){"#### AbsPQN"}`
`r if(do_abspqn){"The data is normalised using Antibody Specific Probabilistic Quotient Normalisation."} `

```{r AbsPQN normalisation between 96 well plates, eval=do_abspqn}

# Retrieve function for AbsPQN
source("/home/.../scripts/norm_AbsPQN.R")

# Make a list with each assay divided into 384-well plates that are divided into 96-well plates, to be able to apply AbsPQN for each 96-well plate
mfi_list_div <- vector("list", length(mfi_list))
names(mfi_list_div) <- names(mfi_list)

for (assay in 1:length(mfi_list_div)) {
  mfi_list_div[[assay]] <- vector("list", 2)
  names(mfi_list_div[[assay]]) <- c("plate384_1", "plate384_2")
  
  for (plate384 in 1:2) {
    mfi_list_div[[assay]][[plate384]] <- vector("list", 4)
    names(mfi_list_div[[assay]][[plate384]]) <- c("plate96_1",
                                                  "plate96_2",
                                                  "plate96_3",
                                                  "plate96_4")
    
    for (plate96 in 1:4) {
      # Empty wells in mfi_list removed here
      mfi_list_div[[assay]][[plate384]][[plate96]] <- mfi_list[[assay]][
        sinfo_w_empty$plate384 == plate384 
        & sinfo_w_empty$plate96 == plate96 
        & sinfo_w_empty$Class != "Empty", ]

      rownames(mfi_list_div[[assay]][[plate384]][[plate96]]) <- rownames(mfi_list[[assay]][
        sinfo_w_empty$plate384 == plate384 
        & sinfo_w_empty$plate96 == plate96 
        & sinfo_w_empty$Class != "Empty", ])
    }
  }
}

# Apply AbsPQN to each 96-well plate, assay-wise
mfi_abspqn <- lapply(mfi_list_div, function(assay) {
  lapply(assay, function(plate384) {
    lapply(plate384, function(plate96) {
      as.data.frame(abspqn(as.matrix(plate96)))
    })
  })
})

# Update mfi dataframe by combining dataframes using the bind_rows function
# Want to update the main mfi dataframe so it can be used in other normalisation steps regardless of which steps have been performed before
# bind_rows function does not retain the row names (and the documentation does not mention any way to do it) so some extra fiddling is done
mfi_list <- lapply(mfi_abspqn, function(assay) {
  br1 <- lapply(assay, function(plate384) {
    row_names <- c(rownames(plate384[[1]]),
                   rownames(plate384[[2]]),
                   rownames(plate384[[3]]),
                   rownames(plate384[[4]]))
    br2 <- bind_rows(plate384)
    rownames(br2) <- row_names
    return(br2)
  })
  row_names <- c(rownames(br1[[1]]),
                 rownames(br1[[2]]))
  br1 <- bind_rows(br1)
  rownames(br1) <- row_names
  return(br1)
})

mfi <- cbind("sample.antibody"=mfi$sample.antibody,
             mfi_list$A3,
             mfi_list$A4,
             mfi_list$A1,
             mfi_list$A2)


```


`r if(do_abspqn){"After AbsPQN the data looks like this"}` 

```{r plot variation after abspqn, eval=do_abspqn, warning=F}

# Do UMAP and plot again to see difference after AbsPQN
umap_norm <- umap_casecontrol(mfi, sinfo)
plot_umap(umap_norm, sinfo[sinfo$Class == "Case" | sinfo$Class == "Control" | sinfo$Class == "Other", ], T, F)$class

# Correlation density after AbsPQN, keep output plot to be able to merge with later plot for report
cordens_abspqn <- cor_dens(mfi, sinfo, xrange=c(0.85, 1), yrange=c(0, 340), plotname="After AbsPQN")
cordens_abspqn

```


`r if(do_ma96){"#### MA for 96-well plate effects"}`
`r if(do_ma96){"The data is normalised using MA for between-plate effects of 96-well plates. The variation looks like this before normalisation."}`

```{r MA normalisation 96 well plates, eval=do_ma96}

# Look at plate variation
plot_96well_box(mfi[, -1], sinfo)

# MA normalisation for between-384plate effects, remove empty wells
mfi_MAnorm <- normn_MA(mD=mfi[, -1], 
                     expGroup=sinfo$plate96)

# Update values in mfi data frame
mfi <- data.frame("sample.antibody"=mfi$sample.antibody,
                  mfi_MAnorm)

```

`r if(do_ma96){"This is what it looks like after the MA normalisation between 96-well plates."}`

```{r box after normalisation, eval=do_ma96}

# Box plots after normalisation
plot_96well_box(mfi[, -1], sinfo, norm=TRUE)

```

Finally, a UMAP is performed again to look at the data after the normalisation. 

```{r UMAP and density on final normalised data, warning=F}

# List of MFI values
mfi_list$A1 <- mfi[, which(binfo1$Assay == "AY485") + 1]
mfi_list$A2 <- mfi[, which(binfo1$Assay == "AY487") + 1]
mfi_list$A3 <- mfi[, which(binfo1$Assay == "AY146") + 1]
mfi_list$A4 <- mfi[, which(binfo1$Assay == "AY483") + 1]

# Do and plot results of UMAP on final normalised data
plot_umap(umap_casecontrol(mfi, sinfo),
          sinfo[sinfo$Class == "Case" | sinfo$Class == "Control" | sinfo$Class == "Other", ], T, F)$class

cor_dens(mfi, sinfo, xrange=c(0.85, 1), yrange=c(0, 340))

```


### Correlation between repeated array

Correlations for antibodies in the repeated array are computed (for plot in antibody_qc.Rmd)

```{r repeat array corr}

# Make sure that antibodies appear in both original and replicate assays
repl_ass_ab <- colnames(mfi)[str_detect(colnames(mfi), "_A1$|_A2$")]

repl_ass_ab <- substr(repl_ass_ab, 1, nchar(repl_ass_ab) - 3) %>% table()

repl_ass_ab <- names(repl_ass_ab[repl_ass_ab == 2])

# All samples
repl_ass_cor <- sapply(repl_ass_ab, function(ab) {
	cor(mfi[, paste0(ab, "_A1")],
			mfi[, paste0(ab, "_A2")],
			method="spearman")
})

# Try without empty samples, pools
no_epr_sampname <- rownames(sinfo %>% filter(Class %in% c("Case", "Control", "Other")))
repl_ass_cor_no_epr <- sapply(repl_ass_ab, function(ab) {
	cor(mfi[no_epr_sampname, paste0(ab, "_A1")],
			mfi[no_epr_sampname, paste0(ab, "_A2")],
			method="spearman")
})

```



